---
title: "BIOST 561: Homework 4"
date: "Due date: 7:00pm on  <u>**Wednesday**</u>, May 29th, 2024"
output: 
  bookdown::html_document2:
    number_sections: false
    toc: true
    toc_float:
      toc_collapsed: true
    theme: united
vignette: >
  %\VignetteIndexEntry{BIOST 561: Homework 4}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(10)
library(testthat)
```

# Instructions

These instructions are essential, so please read them all carefully. 

- Submit your homework on your GitHub page as the RMarkdown (`.Rmd`) and HTML files. 

- Please answer the question prompt and show your code (inline). That is,
all your code should be visible in the knitted chunks.

- To complete this homework, you may write in the `HW3.Rmd` file. 
(It is recommended to complete this homework in R Studio, where clicking the
`Knit` button would knit your homework.)

# Disclosures

- Please disclose below who you collaborated with and if you've used ChatGPT (or any
comparable AI tool) to what extent when completing this homework. See the syllabus
for the course expectations.

# Note about the homework

**Note**: Most of your R code for this homework will **not** be in this homework.
You mainly write `.R` files inside the `R` or `tests` folders.
Therefore, you only need to show a little code inside this R Markdown file.
You only need to write things inside this R Markdown file in the questions that explicitly ask you to do so.

Please use [https://docs.google.com/document/d/1-ZTDPp39zbhsKbQ2FfBoCtiXKQj0jQSg75MAatYfiGI/edit?usp=sharing](https://docs.google.com/document/d/1-ZTDPp39zbhsKbQ2FfBoCtiXKQj0jQSg75MAatYfiGI/edit?usp=sharing) as a resource for this homework.

# Code of conduct for this homework

<mark><u>**Please be respectful to your peers**</u></mark> for this homework. 
Each student's `compute_maximal_partial_clique()` function is anonymized and
being used by every other student for this homework. 
Some implementations of this function work smoother than others. 
Nonetheless, 1) please be respectful to all the implementations throughout this homework, as you might not realize you're talking to the author of an implementation, and 2) please do not feel embarrassed if you (knowing which implementation is yours, since you wrote it) does not work as well as you initially thought. Please do not try to "figure out" which student wrote which implementation.

I am <u>**overwhelmingly confident**</u> that you all wrote an implementation that suits your current comfort in R (which, of course, is different from student to student).

# Q1: Doing a code review of two other implementations

<u>**Question 1B**</u>: Looking at two implementations you were given, summarize what you think the implementations are doing in one to four sentences each. (This is to practice reading other people's code.)

#### First Implementation: `compute_maximal_partial_clique19`
The `compute_maximal_partial_clique19` function seeks to find the largest partial clique within an adjacency matrix that meets a specified edge density threshold, alpha. It uses a greedy approach by first ordering nodes based on their degree and then iteratively building cliques starting from the highest degree nodes. At each step, it adds nodes to the current clique if the resulting submatrix meets the required edge density. The function continues this process to find and return the clique with the highest density that meets the alpha threshold. This method prioritizes nodes with higher connectivity to quickly build dense cliques.

#### Second Implementation: `compute_maximal_partial_clique23`
The compute_maximal_partial_clique23 function aims to identify the largest partial clique in a given adjacency matrix that meets a specified edge density, alpha. The function first verifies the validity of the input matrix and alpha using various checks. It then uses a backtracking approach to explore potential cliques, starting from an empty set and iteratively adding nodes. During this process, it ensures that the current subset of nodes maintains the required edge density. The largest valid clique found is returned along with its edge density.

<u>**Question 1C**</u>
### Recommendations for Improving Code Clarity

#### First Implementation: `compute_maximal_partial_clique19`

1. **Documentation**: The function could benefit from more detailed comments explaining the logic within the loops. For example, explaining why `degrees` are used to order the nodes and what the `possible_edges` calculation represents would help others understand the thought process behind the approach.

2. **Modularization**: Similar to the first implementation, breaking down the main function into smaller helper functions could improve readability and maintainability. For instance, the part of the code that calculates `new_density` could be a separate function, enhancing clarity and enabling easier debugging.

#### Second Implementation: `compute_maximal_partial_clique23`

1. **Variable Naming**: The function uses variables such as `curr_vertex`, `curr_clique`, and `next_vertex` within the `find_maximal_clique` function. These names are clear, but could be more descriptive. For instance, `curr_vertex` could be renamed to `current_node` and `curr_clique` to `current_clique_nodes` to improve readability.

2. **Function Factorization**: The function `is_partial_clique` is nested within the main function. It might be beneficial to move this out as a standalone helper function. This separation can make the code modular and easier to test individual components.

<u>**Question 1D**</u>: 

#### First Implementation: `compute_maximal_partial_clique19`
```{r, error = TRUE}
library(testthat)
test_that("returns empty for an empty clique", {
  res <- generate_partial_clique(n = 5, clique_fraction = 0, clique_edge_density = 0.5)
  result <- compute_maximal_partial_clique19(res$adj_mat, alpha = 0.5)
  expect_equal(result$clique_idx, integer(0))
  expect_equal(result$edge_density, 0)
})

test_that("identifies a full clique", {
  res <- generate_partial_clique(n = 5, clique_fraction = 1, clique_edge_density = 1)
  result <- compute_maximal_partial_clique19(res$adj_mat, alpha = 0.5)
  expect_equal(sort(result$clique_idx), 1:5)
  expect_equal(result$edge_density, 1)
})

test_that("finds clique at exact threshold", {
  res <- generate_partial_clique(n = 5, clique_fraction = 0.5, clique_edge_density = 1)
  result <- compute_maximal_partial_clique19(res$adj_mat, alpha = 0.5)
  expect_true(length(result$clique_idx) >= 2)  # Assuming clique of at least 2 nodes
  expect_true(result$edge_density >= 0.5)
})

test_that("identifies the correct clique among multiple possibilities", {
  n <- 10
  adj_mat <- matrix(0, nrow = n, ncol = n)
  diag(adj_mat) <- 1  # Diagonal elements should be 1

  # First clique: indices 1 to 5, intended lower density (won't meet alpha = 0.7)
  for (i in 1:5) {
    for (j in i:5) {
      if (i != j && runif(1) < 0.4) {  # Ensure i != j to skip diagonal, set lower density
        adj_mat[i, j] <- 1
        adj_mat[j, i] <- 1
      }
    }
  }

  # Second clique: indices 6 to 10, higher density (meets alpha = 0.7)
  for (i in 6:10) {
    for (j in i:10) {
      if (i != j && runif(1) < 0.8) {  # Ensure i != j to skip diagonal, set higher density
        adj_mat[i, j] <- 1
        adj_mat[j, i] <- 1
      }
    }
  }

  result <- compute_maximal_partial_clique19(adj_mat, alpha = 0.7)
  # Expect that the second clique is identified
  expect_true(all(result$clique_idx %in% 6:10))
  expect_true(result$edge_density >= 0.7)
})



test_that("performance on larger graphs", {
  res <- generate_partial_clique(n = 20, clique_fraction = 0.5, clique_edge_density = 0.5)
  system.time({
    result <- compute_maximal_partial_clique19(res$adj_mat, alpha = 0.5)
  }) -> time_taken
  expect_true(time_taken["elapsed"] < 10)  # Test should complete in less than 10 seconds
})
```


#### Second Implementation: `compute_maximal_partial_clique23`
```{r, error = TRUE}
library(testthat)
test_that("returns empty for an empty clique", {
  res <- generate_partial_clique(n = 5, clique_fraction = 0, clique_edge_density = 0.5)
  result <- compute_maximal_partial_clique23(res$adj_mat, alpha = 0.5)
  expect_equal(result$clique_idx, integer(0))
  expect_equal(result$edge_density, 0)
})

test_that("identifies a full clique", {
  res <- generate_partial_clique(n = 5, clique_fraction = 1, clique_edge_density = 1)
  result <- compute_maximal_partial_clique23(res$adj_mat, alpha = 0.5)
  expect_equal(sort(result$clique_idx), 1:5)
  expect_equal(result$edge_density, 1)
})

test_that("finds clique at exact threshold", {
  res <- generate_partial_clique(n = 5, clique_fraction = 0.5, clique_edge_density = 1)
  result <- compute_maximal_partial_clique23(res$adj_mat, alpha = 0.5)
  expect_true(length(result$clique_idx) >= 2)  # Assuming clique of at least 2 nodes
  expect_true(result$edge_density >= 0.5)
})

test_that("identifies the correct clique among multiple possibilities", {
  n <- 10
  adj_mat <- matrix(0, nrow = n, ncol = n)
  diag(adj_mat) <- 1  # Diagonal elements should be 1

  # First clique: indices 1 to 5, intended lower density (won't meet alpha = 0.7)
  for (i in 1:5) {
    for (j in i:5) {
      if (i != j && runif(1) < 0.4) {  # Ensure i != j to skip diagonal, set lower density
        adj_mat[i, j] <- 1
        adj_mat[j, i] <- 1
      }
    }
  }

  # Second clique: indices 6 to 10, higher density (meets alpha = 0.7)
  for (i in 6:10) {
    for (j in i:10) {
      if (i != j && runif(1) < 0.8) {  # Ensure i != j to skip diagonal, set higher density
        adj_mat[i, j] <- 1
        adj_mat[j, i] <- 1
      }
    }
  }

  result <- compute_maximal_partial_clique23(adj_mat, alpha = 0.7)
  # Expect that the second clique is identified
  expect_true(all(result$clique_idx %in% 6:10))
  expect_true(result$edge_density >= 0.7)
})



test_that("performance on larger graphs", {
  res <- generate_partial_clique(n = 20, clique_fraction = 0.5, clique_edge_density = 0.5)
  system.time({
    result <- compute_maximal_partial_clique23(res$adj_mat, alpha = 0.5)
  }) -> time_taken
  expect_true(time_taken["elapsed"] < 10)  # Test should complete in less than 10 seconds
})
```

# Q2: Constructing a simulation study

<mark>**Intent**: The intent of this question is to construct a simulation study that you'll run on Bayes.</mark>

In this question, you'll be designing a simulation. You will be doing this on Bayes for a couple of reasons: 1) To give you experience running code on Bayes, and 2) because your simulations might take a couple of hours to run.

Note: inside the file `compute_maximal_partial_clique_master.R` (which should now be in your `R` folder), I've provided you two functions: `compute_maximal_partial_clique_master()` and `compute_correct_density()`. You will find both functions useful for Q2 and Q3.

<u>**Question 2A**</u>: Please run the following code. If you cannot run it, something has gone wrong with your `UWBiost561` package. (For instance, it might be because your `generate_partial_clique()` function does not work or is not correctly located in your R package.)

```{r, eval = FALSE}
library(UWBiost561)
set.seed(10)
data <- UWBiost561::generate_partial_clique(n = 10, 
                                            clique_fraction = 0.5,
                                            clique_edge_density = 0.95)

set.seed(10)
result1 <- UWBiost561::compute_maximal_partial_clique_master(
  adj_mat = data$adj_mat, 
  alpha = 0.95, 
  number = 11, 
  time_limit = 10
)
result1

set.seed(10)
result2 <- UWBiost561::compute_maximal_partial_clique11(
  adj_mat = data$adj_mat, 
  alpha = 0.95
)
result2
```

As you can see, the `compute_maximal_partial_clique_master()` function: 1) takes a `number` argument (which allows you to control which implementation of `compute_maximal_partial_clique` you're using) and 2) sets a timer (here, of 30 seconds)
so that the function terminates in at most 30 seconds.

Using the provided code as a framework, use the `compute_maximal_partial_clique_master()` with a setting of `time_limit` such that the function terminates prematurely, i.e., `time_limit` is set to be a small number. (That is, if, hypothetically, the function required 40 seconds to complete, but you set `time_limit` to only be 30 seconds.) By changing the ' number ' argument, you can use any implementation of `compute_maximal_partial_clique` you prefer for this question.

Note: For this question, you want to generate an adjacency matrix where `n` is large (i.e., a value close to 50), but `time_limit` is very small (i.e., just a few seconds). You deliberately want to cause a `timed_out` status.

<u>**Question 2B**</u>: 
I am going to use the demo simulation showed during Lecture 8.

<u>**Question 2C**</u>: In your `R` folder, design a function that executes your intended simulation plan. (I am purposely being loose and very open-ended about designing this simulation regarding the inputs or outputs. In contrast to HW3, where everything was spelled out explicitly, I am now giving you the task of meaningfully designing the inputs/outputs.)

Note 1: You want to use `compute_maximal_partial_clique_master()` in your simulation. This would make your life easier when switching between different implementations. The main difficulties in this question are: 

- 1) How to extract and store the results of `compute_maximal_partial_clique_master()`
- 2) How do you decide what meaningful outputs your simulation function should return (for example, you probably don't need to return the `adj_mat` adjacency matrices themselves)? This will highly depend on what your simulation plan is trying to study!
- 3) How to handle the instances when `compute_maximal_partial_clique_master()`: A) times out, B) errors, or C) outputs an alleged `clique_idx` that is invalid because it doesn't form a partial clique with edge density `alpha` or larger.

Note 2: Be aware -- you should be a healthy skeptic when using other people's code in a simulation. Just because a method claims it found a partial clique, you should verify that it is indeed a partial clique with edge density larger than `alpha`. You might want to use the `compute_correct_density()` function provided in `compute_maximal_partial_clique_master.R`.

Note 3: I am asking you to make a <u>**function**</u> to perform your simulation (instead of just an R script) since I will be asking you to test your simulation function in Question 2D below.

Note 4: You should allow your simulation function to take as input the different levels of your simulation study and the number of trials (and other parameters of your choosing). That way, you can more easily "test" a "simple" simulation study in Question 2D below.

Note 5: You (ideally) want to set up your simulation such that for each trial in a level, you generate <u>**one**</u> random adjacency matrix that is used in <u>**all 25**</u> different implementations of `compute_maximal_partial_clique`. This gives a "fair" comparison across all 25 methods. You also might want to set the random seed (via `set.seed()`) before generating that random adjacency matrix for this trial so you can easily reproduce the results from a particular trial if needed.

Note 6: This question is the hardest part of HW4. Since this question is purposely more open-ended, it is harder to use ChatGPT to help you with this function. However, most of your future coding experiences will be open-ended, so practicing "coding with ChatGPT" in more unstructured settings is good. I encourage you to work with your classmates on this question (but, as per our syllabus rules, do not directly copy your classmate's code).

<span style="color: blue;">(There is nothing to report for this question. Your code will be in the `R` folder, **not**
in this R Markdown file.)</span>

<u>**Question 2D**</u>: Create a few (more than one) unit tests for your simulation function. This unit test (like your unit tests in HW3) will be in your `tests/testthat` folder. I am also purposely vague about how many tests or what kind of unit tests to write. This is for you to decide! After all, you're about to unleash your code to perform a (potentially long) simulation test, so you hope you've tested your simulation code well enough for this to be a good use of time.

Note: Your unit tests should be <u>**fast**</u> (i.e., take no more than a minute to run). This means your unit tests should <u>**not**</u> be performing your complete simulation study. After all, it'll be a useless unit test if it takes an hour to figure out if your unit test passed. However, you want your unit tests (which take less than a minute) to give you confidence that your code will work when running the complete simulation study (which might take more than an hour).

<span style="color: blue;">(There is nothing to report for this question. Your code will be in the `tests/testthat` folder, **not**
in this R Markdown file.)</span>

<u>**Question 2E**</u>: Similar to what you did in HW3, please include a screenshot of your output after running `devtools::check()`. You should still have a `UWBiost561` package
that passes all the checks and your unit tests.
(As a guideline, one screenshot should show the first 20-or-so lines of your 
`devtools::check()` results, and a second screenshot should show the last 
20-or-so lines of your results.) You can use the `knitr::include_graphics()` function to include
figures inside this R Markdown file.

The intent of this question is to provide "evidence" that your `devtools::check()`
went smoothly. You do not need to worry about what your screenshots show specifically.

# Q3: Performing the simulation study

<mark>**Intent**: The intent of this question is to perform the simulation study and to practice visualizing simulation results.</mark>

Now that you've written a function to perform your simulation study, you want to write more code to perform the actual simulation study. This will technically involve at least 3 new files in your `vignettes` folder: one `.R` file to perform the simulation study, one accompanying `.slurm` script to submit those above `.R` file to Bayes, and one `.R` file to load the simulation results and visualize the results. The following questions will guide you through these three files.

<u>**Question 3A**</u>: In your `vignettes` folder, make a file called `HW4_simulation_execute.R`, which will be a R script that: 1) loads your `UWBiost561` package, and 2) executes your simulation function inside your `UWBiost561` package.
This simulation will be the "full" simulation study you outlined in Question 2B above.
(This is the specific `.R` file that might take a couple of hours to finish, depending on how complex your simulation study is.) The end of your `HW4_simulation_execute.R` should save your simulation results as `HW4_simulation.RData`, an `.RData` file. (For example, the argument to the `save()` function could be `file = ~/HW4_simulation.RData`. This would save the results to a `HW4_simulation.RData` file under your home directory on Bayes.) 

Then, make an accompanying `HW4_simulation_execute.slurm` that submits your `HW4_simulation_execute.R` as a job to Bayes. Please limit yourself to `10gb` of memory (i.e., a line in `HW4_simulation_execute.slurm` should read `#SBATCH --mem-per-cpu=10gb`). (Your simulation study should not require more than 10 Gigabytes of memory. If it does, you are likely not being practical in your simulation study.)

<span style="color: blue;">(There is nothing to report for this question. Your code will be in the `vignettes` folder, **not**
in this R Markdown file.)</span>

See Lectures 7 and 8 for more details on how to do this.

<u>**Question 3B**</u>: Now, install both `igraph` and your `UWBiost561` package in R.
The `igraph` package can be installed as usual (i.e., via `install.packages(igraph)` in an interactive R session on Bayes). Lecture 8 and [https://docs.google.com/document/d/1-ZTDPp39zbhsKbQ2FfBoCtiXKQj0jQSg75MAatYfiGI/edit?usp=sharing](https://docs.google.com/document/d/1-ZTDPp39zbhsKbQ2FfBoCtiXKQj0jQSg75MAatYfiGI/edit?usp=sharing) will give you a few options on how to install your `UWBiost561` package on Bayes.

<span style="color: blue;">(There is nothing to report for this question.)</span>

<u>**Question 3C**</u>: Now, run your simulation study. In the terminal, this means you will 1) navigate to your `UWBiost561/vignettes` folder on Bayes (this might be something like `cd ~/UWBiost561/vignettes`) and 2) run the command: `sbatch HW4_simulation_execute.slurm`.

Note 1: This question might cause you pain since you might not know if your script is working as intended. (Additionally, suppose you did not test your simulation function thoroughly. In that case, your `HW4_simulation_execute.slurm` script might crash, and you'll be forced to debug your `UWBiost561` package and re-install it, etc. It's a pain. Trust me, you want to test your simulation function thoroughly before you do this Question.)

<mark><u>**Note 2:**</u> Do <u>**not**</u> parallelize your simulation across multiple cores or write a batch SLURM script to submit multiple jobs for this question. (Many of you will likely be working on the homework at similar times, and I cannot guarantee that the server can allow an entire course of students to submit parallelized jobs simultaneously.) If this note does not make any sense to you, don't worry about it.</mark>

<span style="color: blue;">(There is nothing to report for this question. You are simply running your `HW4_simulation_execute.slurm` file.)</span>

<u>**Question 3D**</u>: Finally, create a file called `HW4_simulation_plot.R` under the `vignettes` folder. You will design this script to load your saved results in `HW4_simulation.RData`, visualize the results, and save the plot into your `vignettes` folder. Name this one plot as `HW4_simulation.png`. (You can save your plot in a different file format, but `.png` is the easiest one to work with.)

How complicated should this script be? It depends on the results you saved in `HW4_simulation.RData`. For example, if you saved the clique size of each method in each trial for each level, then you might need to compute the average clique size across all the trials for each level in this script. 

What should you plot? It depends on what your simulation plan was intended to study (which you wrote in Question 2B). I'm leaving this open-ended so you can practice determining how best to demonstrate the results you're trying to study. 

Note 1: You can run your `HW4_simulation_plot.R` script <u>**interactively**</u> on Bayes. You do not need to write a `.slurm` script for this if you prefer not to (since your entire `HW4_simulation_plot.R` script should take only a few minutes to run. After all, you're simply loading in and plotting the results. There shouldn't be any very fancy computation being performed in this script, and indeed, you shouldn't be computing any maximal partial cliques in this script).

Note 2: Feel free to consult [https://r-graph-gallery.com/](https://r-graph-gallery.com/) to get ideas on how to meaningfully visualize your results.

<span style="color: blue;">(There is nothing to report for this question. Your code will be in the `vignettes` folder, **not**
in this R Markdown file.)</span>

<u>**Question 3E**</u>: Now, on Bayes, there should be at least four new files in the `vignettes` folder due to Question 3A-3D: `HW4_simulation_execute.R`, `HW4_simulation_execute.slurm`, `HW4_simulation_plot.R`, and `HW4_simulation.png`. (There might be many other files, but I am okay with it if there are more files than needed.) Commit and push all these files via Git (using the command line on Bayes) onto GitHub.com (via `git push`), and then pull all these files via Git onto your local laptop. (You can pull either through the command line via `git pull` or the RStudio GUI.)

Finally, include your plot in this R Markdown file. (I'm having you put your plot in your `vignettes` folder and push/pull your `UWBiost561` package because I'm assuming you knit this `HW4.Rmd` file locally on your laptop, but you need to somehow get your plot from Bayes onto your local laptop.)

# Q4: Describing your final project

<mark>**Intent**: The intent of this question is to make sure you have a plan for the final project.</mark>

I will release the final project specifications (again, which involve making a PkgDown website of any R package of your choosing of any scope) on Canvas. This will be released (at the latest) by May 19th.

Please write (in one to five sentences) what you are thinking of doing for your final project. If your answer hasn't changed since HW3, you can copy-paste your answer from HW3 into this question for HW4.

# Q5: Feedback (Optional)

This "question" is an additional way for students to communicate with instructors. 
You could include positive feedback about topics you enjoyed learning in this module, critiques about 
the course difficulty/pacing, 
or some questions/confusions you had about course material.
Your feedback can help shape the course for the rest of this quarter and 
future years. Please be mindful and polite when providing feedback.
You may leave this question blank.


